### 2017-01-17 responses/response.tex

\noindent
Contextual dependencies can often be satisfied in multiple ways\footnote{For example, see the
discussion about distributed programming in Section~3.2.2 of the thesis.}. By capturing the
contextual properties in the type-and-coeffect system, we are making the existing
run-time ambiguities \emph{more explicit}. Thus the proposal is to focus on systems with
\emph{coeffect-directed semantics} rather than modifying the calculi that inspired the work
in the thesis in order to force coherence.

\section{Resolving ambiguity}
\label{sec:ambiguity}
\vspace{-1em}

The case where the semantics of a context-aware program depends on the typing derivation is best
demonstrated using the system for implicit parameters (Section~3.2.1 of the thesis). Consider the
following example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f} = (~\kvd{let}~\ident{?x} = 10~\kvd{in}~
  \lambda \ident{v} \rightarrow \ident{?x})~\kvd{in}\\[-0.2em]
\kvd{let}~\ident{?x} = 15~\kvd{in}~\ident{f}~0
\end{array}
\end{equation*}
%
Here, a function value $\lambda \ident{v} \rightarrow \ident{?x}$ is created in a context where the
implicit parameter $\ident{?x}$ has a value 10. The function is then called in a context where
$\ident{?x}$ has a value 15. Is the result of the expression $10$ or $15$? The choice of the
run-time semantics needs to be reflected in the type system.

\paragraph{Resolution in Haskell.}
In Haskell, the implicit parameters from the \emph{declaration site} are preferred and so the result
is 10. Correspondingly, the type system only permits one typing -- with the parameter \ident{?x}
required on the declaration-side (allowing no rebinding at the call site). However, if the function
value was defined in a context not containing \ident{?x}, the value from the call site would be used.
Thus, Haskell uses a \emph{defaulting mechanism} for implicit parameters, similar to that of F\#
operators discussed in Section~\ref{sec:background-practice}.

@@

The type systems guarantee that \emph{coeffect indices} will match in the resulting 
translation. The relationship from indices to values is injective in the general case, but 
we can also show a number of important properties (which will be clarified in the revised
version) for the concrete systems, \eg~an upper bound on the number of required
past values in causal dataflow.

% ==================================================================================================

\section{Summary}

An interesting aspect of many context-aware computations is that they require access to some
context, but do not explicitly specify where the context comes from. In case of implicit parameters,
the values can be provided either by the call site or by the declaration site.

### 2017-01-17 main.tex
Edits pages 31-94

  The type system is complemented with a translational semantics for coeffect-based context-aware
  programming languages (Chapter~\ref{ch:semantics}). The semantics is inspired by a categorical
  model based on \emph{indexed comonads} and it translates source context-aware program into a
  target program in a simple functional language with comonadically-inspired primitives. We give
  concrete definition of the primitives for dataflow, implicit parameters and liveness and
  present a syntactic safety proof for these three languages.

  The following page provides a detailed overview of the content of Chapters~\ref{ch:flat} and
  Chapters~\ref{ch:semantics}, highlighting the split between general definitions and properties
  (about the coeffect calculus) and concrete definitions and properties (about concrete
  context-awre language). The Chapter~\ref{ch:structural} mirrors the same development for
  \emph{structural coeffect systems}.
}
\part{Coeffect calculi}
\label{part:coeffect-calculi}

@@

\begin{center}
\begin{tabular}{ | l | l | l |}
\hline
\multicolumn{3}{ | c | }{\spacedlowsmallcaps{Chapter 5}} \\ \hline \hline
& \textsc{Coeffect calculus} & \textsc{Language-specific} \\ \hline
\textsc{Categorical}
  & \hspace{-0.5em}\begin{tabular}{l} Indexed comonads \\[-0.3em] (Sectiuon~\ref{sec:semantics-flat-idx}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Examples including indexed \\[-0.3em] product, list and maybe \\[-0.3em] comonads (Section~\ref{sec:semantics-flat-monoidal}) \end{tabular} \\
\hline
  & \hspace{-0.5em}\begin{tabular}{l} Categorical semantics \\[-0.4em] of coeffect $\lambda$-calculus \\[-0.3em] (Section~\ref{sec:semantics-flat-calculus}) \end{tabular}
  & \\ \hline
\textsc{Translational}
  & \hspace{-0.5em}\begin{tabular}{l} Functional target \\[-0.3em] language (Section~\ref{sec:semantics-translation-target}) \end{tabular}
  &  \\
\hline
  & \hspace{-0.5em}\begin{tabular}{l} Translation from coeffect \\[-0.3em] $\lambda$-calculus to target \\[-0.3em]  language (Section~\ref{sec:semantics-translation-transl}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Translation for language-\\[-0.3em] specific extensions (\kvd{prev}, \ident{?p}) \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\ \hline
\textsc{Operational}
  & \hspace{-0.5em}\begin{tabular}{l} Abstract comonadically- \\[-0.3em] inspired primitives \\[-0.3em] (Section~\ref{sec:semantics-translation-transl}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Concrete reduction rules for \\[-0.3em] comonadically-inspired primitives \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\
\hline
  &
  & \hspace{-0.5em}\begin{tabular}{l} Reduction rules for language-\\[-0.3em] specific extensions (\kvd{prev}, \ident{?p}) \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\
\hline
  & \hspace{-0.5em}\begin{tabular}{l} Sketch of generalized \\[-0.3em] syntactic soundness \\[-0.3em] (Section~\ref{sec:semantics-generalising}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Syntactic soundness \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\ \hline
\end{tabular}
\end{center}

>>>>

  The type system is complemented with a translational semantics for coeffect-based context-aware
  programming languages (Chapter~\ref{ch:semantics}). The semantics is inspired by a categorical
  model based on \emph{indexed comonads} and it translates a source context-aware program into a
  target program in a simple functional language with comonadically-inspired primitives. We give a
  concrete definition of the primitives for dataflow, implicit parameters and liveness and
  present a syntactic safety proof for these three languages.

  The following page provides a detailed overview of the content of Chapters~\ref{ch:flat} and
  Chapters~\ref{ch:semantics}, highlighting the split between general definitions and properties
  (about the coeffect calculus) and concrete definitions and properties (about concrete
  context-awre language). Chapter~\ref{ch:structural} mirrors the same development for
  \emph{structural coeffect systems}.
}
\part{Coeffect calculi}
\label{part:coeffect-calculi}

@@

\begin{center}
\begin{tabular}{ | l | l | l |}
\hline
\multicolumn{3}{ | c | }{\spacedlowsmallcaps{Chapter 5}} \\ \hline \hline
& \textsc{Coeffect calculus} & \textsc{Language-specific} \\ \hline
\textsc{Categorical}
  & \hspace{-0.5em}\begin{tabular}{l} Indexed comonads \\[-0.3em] (Section~\ref{sec:semantics-flat-idx}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Examples including indexed \\[-0.3em] product, list and maybe \\[-0.3em] comonads (Section~\ref{sec:semantics-flat-monoidal}) \end{tabular} \\
\hline
  & \hspace{-0.5em}\begin{tabular}{l} Categorical semantics \\[-0.4em] of coeffect $\lambda$-calculus \\[-0.3em] (Section~\ref{sec:semantics-flat-calculus}) \end{tabular}
  & \\ \hline
\textsc{Translational}
  & \hspace{-0.5em}\begin{tabular}{l} Functional target \\[-0.3em] language (Section~\ref{sec:semantics-translation-target}) \end{tabular}
  &  \\
\hline
  & \hspace{-0.5em}\begin{tabular}{l} Translation from coeffect \\[-0.3em] $\lambda$-calculus to target \\[-0.3em]  language (Section~\ref{sec:semantics-translation-transl}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Translation for language-\\[-0.3em] specific extensions (\kvd{prev}, \ident{?p}) \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\ \hline
\textsc{Operational}
  & \hspace{-0.5em}\begin{tabular}{l} Abstract comonadically- \\[-0.3em] inspired primitives \\[-0.3em] (Section~\ref{sec:semantics-translation-transl}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Concrete reduction rules for \\[-0.3em] comonadically-inspired primitives \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\
\hline
  &
  & \hspace{-0.5em}\begin{tabular}{l} Reduction rules for language-\\[-0.3em] specific extensions (\kvd{prev}, \ident{?p}) \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\
\hline
  & \hspace{-0.5em}\begin{tabular}{l} Sketch of generalized \\[-0.3em] syntactic soundness \\[-0.3em] (Section~\ref{sec:semantics-generalising}) \end{tabular}
  & \hspace{-0.5em}\begin{tabular}{l} Syntactic soundness \\[-0.3em] (Sections~\ref{sec:semantics-proofs-df} and \ref{sec:semantics-proofs-impl}) \end{tabular} \\ \hline
\end{tabular}
\end{center}

### 2017-01-17 text/Chapter02.tex
Edits pages 31-94

\noindent
When we view variable access as a context requirement, we can see that the body of the function
($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
\ident{y}. In static binding, the context demands of the body can be placed on the scope in
which the function is defined (declaration-site). In dynamic binding, the requirements are
\emph{delayed} and are placed on the scope in which the function is called (call site).

@@

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for Haskell's implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in static scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call site.
In the example on the left, \ident{?x} is bound to $10$ and so the function \ident{f} has no
delayed context demands and thus the expression evaluates to $10$. On the right,
the context demand for \ident{?x} is \emph{delayed} and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

@@

In this thesis, we consider various notions of context, using implicit parameters as one
of several motivating examples. Implicit parameters are a particularly valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
a function can be satisfied using the context available at declaration-site or using the context
available at the call site. Recall our earlier printer example -- a language that
provides access to resources in the context needs to provide enough flexibility in handling such
ambiguities, be it implicit parameter values available in scope or printers available in the
physical environment.

We aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context demands are split between the declaration-site and the call site. While
Haskell's solution for implicit parameters might be the most reasonable one for that particular
case, this thesis argues that other notions of context require different domain-specific choices
and the general framework of context-aware programming should make that possible.

>>>>

\noindent
When we view variable access as a context requirement, we can see that the body of the function
($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
\ident{y}. In static binding, the context demands of the body can be placed on the scope in
which the function is defined (declaration site). In dynamic binding, the requirements are
\emph{delayed} and are placed on the scope in which the function is called (call site).

@@

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for Haskell's implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in static scope, then the value
is statically bound and the context requirement is satisfied using the declaration site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call site.
In the example on the left, \ident{?x} is bound to $10$ and so the function \ident{f} has no
delayed context demands and thus the expression evaluates to $10$. On the right,
the context demand for \ident{?x} is \emph{delayed} and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

@@

In this thesis, we consider various notions of context, using implicit parameters as one
of several motivating examples. Implicit parameters are a particularly valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
a function can be satisfied using the context available at declaration site or using the context
available at the call site. Recall our earlier printer example -- a language that
provides access to resources in the context needs to provide enough flexibility in handling such
ambiguities, be it implicit parameter values available in scope or printers available in the
physical environment.

We aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context demands are split between the declaration site and the call site. While
Haskell's solution for implicit parameters might be the most reasonable one for that particular
case, this thesis argues that other notions of context require different domain-specific choices
and the general framework of context-aware programming should make that possible.
